Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-02-01 MScDiplom_Lemenkova.tex
added Index

\documentclass[10pt, a4paper]{article}
\usepackage{color}
\usepackage{setspace}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage[a4paper=true, ]{hyperref}

@@

\section*{}
	\markright{Table of Content}
	\addcontentsline{toc}{section}{Table of Content}
	\tableofcontents 
\pagebreak

@@

\section*{Document Outline}
	\markright{Structure of the MSc Thesis}
	\addcontentsline{toc}{section}{Document Outline}
\textit{Chapter 1}: \nameref{sec:1} is an introductory section. It describes general background of the research problem, outlines the need and actuality of the problem (including the environmental vulnerability of the seagrasses), and highlights limitations and possibilities of the remote sensing application for the current work. In this chapter we also set up research objectives, put research questions for hypothesis testing and sketch research approach for the proposed work.\\
In \textit{Chapter 2}: \nameref{sec:2} we review the existing literature and reported research experience on the similar problem: studies of spectral separability of various seafloor types, limitations and advantages of the remote sensing techniques applied for seagrass mapping seagrass ecology and mapping seagrass environment. We considered not only the Mediterranean environment, but also papers from Australian and Chinese scientists, because seagrass monitoring is most actively developed in the southern regions of our planet. The review of the existing RTM algorithms for the retrieval of the optical parameters, as well as description of various tools for the spectrometric measurements - are given in the same Chapter as well. More close attention has been given to the imaging hyperspectral radiometer Trios RAMSES, used for the data collection.\\
\textit{Chapter 3}: \nameref{sec:3} deals with the methods and materials used in current work. We start our discussion from the fieldwork area location and describe tools and instruments used during the fieldwork, as well as sampling design. Procedures of the data pre-processing and capturing imagery from the Google Earth are also presented in the same chapter. \\
\textit{Chapter 4}: \nameref{sec:4} presents to the reader the main results of the current MSc work, obtained during the processing fieldwork and other collected data. It starts from the review of the collected data from different sources, then describes the modelling component of the research, namely WASI colour simulation and plotted resulting graphs. We also discuss here the particularities of various sensors, suitable enough for deriving radiometric information to discriminate sand from the seagrass. Furthermore, this chapter includes analysis of the spectral signatures of various seafloor cover types in conditions of changing environment, possibility and limitations for \textit{P.oceanica} spectral discrimination, statistical analysis of the data sets and mapping based on images processing.\\
\textit{Chapter 5}: \nameref{sec:5} summarizes and briefly discusses again the main results, reported in detail in a previous chapter. This chapter proposes some final discussions to the reader, and comes to the main research outcomes.\\
Finally, we suggest some recommendations for any further research focusing on the seagrass environment along the coasts of Crete (or other Mediterranean areas) in the final \nameref{sec:6}.\\
A lot of plotted graphs, statistical outcomes, tables, auxiliary (yet relevant to our research) pictures and images, illustrating our work - are collected in the \nameref{app}. Due to the standard editorial limitations of the current MSc work, it was not possible to include all them to the main chapters. However, we tried to make the structure of the Appendices most clear and easy-readable as possible, by dividing the Appendices into 10 various sub-sections and referencing to them from the main text where necessary. \\
The \nameref{bib} section includes literature and internet resources used for the current work. The cross-referencing and web hyper-referencing are used in the whole document to make reading more quick, effective and informative.

@@

\pagebreak
\singlespace
\section*{List of Tools}\label{sec:0}
	\markright{List of Tools}
	\addcontentsline{toc}{section}{List of Tools}

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to the marine environment, they are however, more closely related to the lily family
(\textit{Liliaceae}) than to the true grass, despite their name \textit{seagrass}, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}. Seagrasses create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

@@

\subsection{Background}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species (Fig. \ref{fig:2}), belonging to two orders
(\textit{Hydrocharitales} and \textit{Najadales}), four families (\textit{Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae}), and 12 genera (\textit{Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera} and \textit{Phyllospandix}) \cite{Kuo89}\label{Kuo89}.

@@

The distribution of the seagrasses is strongly influenced by several environmental factors, which
include climate (mostly, tropical and temperate areas), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutrient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species in Mediterranean area \cite{Borum04}\label{Borum04}: \textit{Zostera marina,
Zostera noltii, Cymodocea nodosa} and \textit{P.oceanica}. In Greece the common species are \textit{P.oceanica}
\textit{(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann} and \textit{Halophila
stipulacea} \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, \textit{Cymodocea nodosa} is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When \textit{P.oceanica} beds regresses, \textit{C.nodosa} often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
\textit{C.nodosa}, and \textit{Z.noltii} do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. One
of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06} \label{Leoni07}.

@@

The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays P.oceanica is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.\\
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.

@@

\subsubsection{General characteristics of \textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows in the Mediterranean between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.
The dominant and most productive coastal ecosystem of the Mediterranean, \textit{P.oceanica} is
spatially restricted to the Mediterranean area (Fig.\ref{fig:5}), with its extension limited by the western part
of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphological \textit{P.oceanica} consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots
and short rhizomes (0.5-2.0 mm). The leaves are, perhaps, the most particular characteristics of

@@

of the meadow health status. \\
There are several environmental factors, determining the growth of the \textit{P.oceanica}. The adaptation
to dry-summer subtropical climate reduces its extension to Mediterranean area only (Fig.\ref{fig:5}).
Besides, the distribution of the seagrasses changes with water depth: it is noticed \cite{Dural10}\label{Dural10} that
the highest flowering density is usually in the 4-7 m depth. 
\textit{P.oceanica} flowers appeared in shallow
stands in September while in November only in stands deeper than 15 m. This time delay is caused by
the different maximum summer temperatures at those depths \cite{Buia91}\label{Buia91}.

@@

The phenology of \textit{P.oceanica} is also affected by the coastal bathymetry: in the isolated meadows in
shallow waters plants have shorter and falciform leaves, compared to ones in the deeper and central
areas \cite{Dural10}\label{Dural10}. In \textit{P.oceanica} flower abundance is related to the structure of the meadow with
the maximal flower density in the densest stands, while the occurrence of flowering is regulated by
environmental factors \cite{Buia91}\label{Buia91}. Phenology of the \textit{P.oceanica} undergoes 
modifications with varying seasons during the year: during the flowering period (ca 3 months long)
the number of leaves on the flowering shoots decrease. Changes of the leaf growth also appear in the
flowering shoots with longer oldest leaves and shorter and narrower leaves induced during the
flowering \cite{Gobert01}\label{Gobert01}.

\subsection{Research problem}
Monitoring of the marine benthic ecosystems of seagrasses is essential for the environmental
assessment of the coastal zones. It increases our knowledge of the seagrass ecology, highlights 
threats to the seagrass and preventing them from possible losses and degrading and improves
techniques and methods of the underwater-based observations. Mapping the seagrass contributes to
the evaluating of the seagrass current distribution, analysis of its dynamics and changes over time, as
well as estimations of the degrading of seagrass meadows for the purpose of the coastal management.\\
Precise, correct and up-to-date information about the distribution of \textit{P.oceanica} is necessary for the
sustainable conservation of the marine environment and ecosystems in Mediterranean area, being an
important contribution to the environmental coastal zone management \cite{Pergent-Martini06}\label{Pergent-Martini06}.
However, mapping the seagrass has limitations due to the specific location and characteristics of the
research object. The remote sensing techniques have traditionally been widely used for the seagrass
monitoring.

@@

\subsection{Research objective}
The current MSc research aims to explore the environmental conditions for the spatial distribution of \textit{P.oceanica} seagrass along the northern coast of Crete Island, based on the remote sensing and GIS techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources:
	\begin{itemize}
		\item[] i) spectra of \textit{P.oceanica}, carbonate sand, silt and other seafloor types
		\item[] ii) satellite imagery: Landsat TM, Landsat ETM+
		\item[] iii) aerial photos: Google Earth
		\item[] iv) \textit{in-situ} fieldwork data of underwater videographic measurements
		\item[] v) vector GIS layers.
	\end{itemize}
The research aims to explore the limitations of the application of the Radiative Transfer Models (further RTMs) and remote sensing techniques 
towards the study of the environmental properties of the \textit{P.oceanica} meadows distribution
over the selected locations of northern Crete. The main research objective is monitoring the seagrass \textit{P.oceanica} in selected areas of Crete and analyzing the environmental conditions for the growth of \textit{P.oceanica}. \\
This research is supported by the \textit{in-situ} measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: spectral modelling by means of the RTM Water Colour Simulator WASI,  underwater videometric measurements made by Olympus camera, aerial Google Earth and satellite images from different sources, spatial GIS and statistical analysis.

\subsubsection{General objective}
The main objective of this study is to analyse the optical properties of the seagrass \textit{P.oceanica} and other seafloor types, and to apply the remote sensing techniques towards the investigation of the seagrass distribution in the selected locations along the northern coast of Crete.
General objectives:
	\begin{itemize}
		\item[*]Analysing spectral reflectance of the \textit{P.oceanica} and other seafloor cover types means of radiative transfer model tools (RTMs), using WASI.
		\item[*]Mapping spatial distribution of the seagrass \textit{P.oceanica} over selected locations along the northern coasts of Crete Island 
	\end{itemize}

\subsubsection{Specific objectives}
	\begin{itemize}	
		\item[]1.	To study narrow-band spectral reflectance properties of \textit{P.oceanica} and other seafloor cover types (sand and silt) using WASI water colour simulation software 
		\item[]2.	To use methods of the in situ diving observations and underwater videometric measurements by Olympus camera in order to receive large-scale imagery of the P.oceanica mattes 
		\item[]3.	To apply remote sensing data (Google Earth aerial images, Landsat TM and ETM+ satellite images) for the monitoring of the seagrass meadows distribution 
		\item[]4.	To perform supervised images classification for the thematic mapping of the \textit{P.oceanica} seagrass distribution along the selected locations over the coasts of northern Crete. 
	\end{itemize}

>>>>

\documentclass[10pt, a4paper]{article}
\usepackage{color}
\usepackage{setspace}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
\usepackage{makeidx}
\makeindex
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage[a4paper=true, ]{hyperref}

@@

\section*{}
	\markright{Table of Content}
	\addcontentsline{toc}{section}{Table of Content}
	\setlength{\textwidth}{6.5in}
	\addtolength{\voffset}{+3pt}
	\textheight = 620pt
	\tableofcontents 
\pagebreak

@@

\section*{Document Outline}
	\markright{Structure of the MSc Thesis}
	\addcontentsline{toc}{section}{Document Outline}
\textit{Chapter 1}: \nameref{sec:1} is an introductory section. It describes general background of the research problem, outlines the need and actuality of the problem (including the environmental vulnerability of the seagrasses), and highlights limitations and possibilities of the remote sensing application for the current work. In this chapter we also set up research objectives, put research questions for hypothesis testing and sketch research approach for the proposed work.\\
In \textit{Chapter 2}: \nameref{sec:2} we review the existing literature and reported research experience on the similar problem: studies of spectral separability of various seafloor types, limitations and advantages of the remote sensing techniques applied for seagrass mapping seagrass ecology and mapping seagrass environment. We considered not only the Mediterranean environment, but also papers from Australian and Chinese scientists, because seagrass monitoring is most actively developed in the southern regions of our planet. The review of the existing RTM algorithms for the retrieval of the optical parameters, as well as description of various tools for the spectrometric measurements - are given in the same Chapter as well. More close attention has been given to the imaging hyperspectral radiometer Trios RAMSES, used for the data collection.\\
\textit{Chapter 3}: \nameref{sec:3} deals with the methods and materials used in current work. We start our discussion from the fieldwork area location and describe tools and instruments used during the fieldwork, as well as sampling design. Procedures of the data pre-processing and capturing imagery from the Google Earth are also presented in the same chapter. \\
\textit{Chapter 4}: \nameref{sec:4} presents to the reader the main results of the current MSc work, obtained during the processing fieldwork and other collected data. It starts from the review of the collected data from different sources, then describes the modelling component of the research, namely WASI colour simulation and plotted resulting graphs. We also discuss here the particularities of various sensors, suitable enough for deriving radiometric information to discriminate sand from the seagrass. Furthermore, this chapter includes analysis of the spectral signatures of various seafloor cover types in conditions of changing environment, possibility and limitations for \textit{P.oceanica} spectral discrimination, statistical analysis of the data sets and mapping based on images processing.\\
\textit{Chapter 5}: \nameref{sec:5} summarizes and briefly discusses again the main results, reported in detail in a previous chapter. This chapter proposes some final discussions to the reader, and comes to the main research outcomes.\\
Finally, we suggest some recommendations for any further research focusing on the seagrass environment along the coasts of Crete (or other Mediterranean areas) in the final \nameref{sec:6}.\\
A lot of plotted graphs, statistical outcomes, tables, auxiliary (yet relevant to our research) pictures and images, illustrating our work - are collected in the \nameref{app}. Due to the standard editorial limitations of the current MSc work, it was not possible to include all them to the main chapters. However, we tried to make the structure of the Appendices most clear and easy-readable as possible, by dividing the Appendices into 10 various sub-sections and referencing to them from the main text where necessary. \\
The \nameref{bib} section includes literature and internet resources used for the current work. The cross-referencing and web hyper-referencing are used in the whole document to make reading more quick, effective and informative.\\
An Index is placed in the end of the Document to help the reader find what he is looking for.

@@

\pagebreak
\singlespace
\section*{List of Tools}\label{sec:0}
	\markright{List of Tools}
	\addcontentsline{toc}{section}{List of Tools}\index{Software used}

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to the marine environment, they are however, more closely related to the lily family
(\textit{Liliaceae}) than to the true grass, despite their name \textit{seagrass}, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}. Seagrasses\index{Seagrasses} create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

@@

\subsection{Background}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species\index{Seagrasses!global species} (Fig. \ref{fig:2}), belonging to two orders
(\textit{Hydrocharitales} and \textit{Najadales}), four families (\textit{Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae}), and 12 genera (\textit{Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera} and \textit{Phyllospandix}) \cite{Kuo89}\label{Kuo89}.

@@

The distribution of the seagrasses is strongly influenced by several environmental factors, which
include climate (mostly, tropical and temperate areas\index{Seagrasses!distribution of}), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutrient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species \index{Seagrasses!Mediterranean species} in Mediterranean area \cite{Borum04}\label{Borum04}: \textit{Zostera marina,
Zostera noltii, Cymodocea nodosa} and \textit{P.oceanica}. In Greece the common species are \textit{P.oceanica}
\textit{(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann} and \textit{Halophila
stipulacea} \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, \textit{Cymodocea nodosa} is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When \textit{P.oceanica} beds regresses, \textit{C.nodosa} often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
\textit{C.nodosa}, and \textit{Z.noltii} do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities\index{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. One
of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06} \label{Leoni07}.

@@

The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays P.oceanica is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution\index{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes\index{Seagrasses!natural impacts on} can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subjectto catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.\\
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.

@@

\subsubsection{General characteristics of \textit{P.oceanica}}\index{Seagrasses!\textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows in the Mediterranean between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.
The dominant and most productive coastal ecosystem of the Mediterranean, \textit{P.oceanica} is
spatially restricted to the Mediterranean area\index{Seagrasses!\textit{P.oceanica}!distribution} (Fig.\ref{fig:5}), with its extension limited by the western part
of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphologically\index{Seagrasses!\textit{P.oceanica}!morphology}, \textit{P.oceanica} consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots
and short rhizomes (0.5-2.0 mm). The leaves are, perhaps, the most particular characteristics of

@@

of the meadow health status. \\
There are several environmental factors, determining the growth of the \textit{P.oceanica}. The adaptation\index{Seagrasses!\textit{P.oceanica}!environmental adaptation}
to dry-summer subtropical climate reduces its extension to Mediterranean area only (Fig.\ref{fig:5}).
Besides, the distribution of the seagrasses changes with water depth: it is noticed \cite{Dural10}\label{Dural10} that
the highest flowering density is usually in the 4-7 m depth. 
\textit{P.oceanica} flowers appeared in shallow
stands in September while in November only in stands deeper than 15 m. This time delay is caused by
the different maximum summer temperatures at those depths \cite{Buia91}\label{Buia91}.

@@

The phenology\index{Seagrasses!\textit{P.oceanica}!phenology} of \textit{P.oceanica} is also affected by the coastal bathymetry: in the isolated meadows in
shallow waters plants have shorter and falciform leaves, compared to ones in the deeper and central
areas \cite{Dural10}\label{Dural10}. In \textit{P.oceanica} flower abundance is related to the structure of the meadow with
the maximal flower density in the densest stands, while the occurrence of flowering is regulated by
environmental factors \cite{Buia91}\label{Buia91}. Phenology of the \textit{P.oceanica} undergoes 
modifications with varying seasons during the year: during the flowering period (ca 3 months long)
the number of leaves on the flowering shoots decrease. Changes of the leaf growth also appear in the
flowering shoots with longer oldest leaves and shorter and narrower leaves induced during the
flowering \cite{Gobert01}\label{Gobert01}.

\subsection{Research problem}\index{Research}
Monitoring\index{Research!problem} of the marine benthic ecosystems of seagrasses is essential for the environmental
assessment of the coastal zones. It increases our knowledge of the seagrass ecology, highlights 
threats to the seagrass and preventing them from possible losses and degrading and improves
techniques and methods of the underwater-based observations. Mapping the seagrass contributes to
the evaluating of the seagrass current distribution, analysis of its dynamics and changes over time, as
well as estimations of the degrading of seagrass meadows for the purpose of the coastal management.\\
Precise, correct and up-to-date information about the distribution of \textit{P.oceanica} is necessary for the
sustainable conservation of the marine environment and ecosystems in Mediterranean area, being an
important contribution to the environmental coastal zone management \cite{Pergent-Martini06}\label{Pergent-Martini06}.
However, mapping the seagrass has limitations due to the specific location and characteristics of the
research object. The remote sensing techniques have traditionally been widely used for the seagrass
monitoring.

@@

\subsection{Research objective}\index{Research!objective}
The current MSc research aims to explore the environmental conditions for the spatial distribution of \textit{P.oceanica} seagrass along the northern coast of Crete Island, based on the remote sensing and GIS techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources \index{Data}:
	\begin{itemize}
		\item[] i) spectra of \textit{P.oceanica}, carbonate sand, silt and other seafloor types
		\item[] ii) satellite imagery: Landsat TM, Landsat ETM+
		\item[] iii) aerial photos: Google Earth
		\item[] iv) \textit{in-situ} fieldwork data of underwater videographic measurements
		\item[] v) vector GIS layers.
	\end{itemize}
The research aims to explore the limitations of the application of the Radiative Transfer Models (further RTMs) and remote sensing techniques 
towards the study of the environmental properties of the \textit{P.oceanica} meadows distribution
over the selected locations of northern Crete. The main research objective is monitoring the seagrass \textit{P.oceanica} in selected areas of Crete and analyzing the environmental conditions for the growth of \textit{P.oceanica}. \\
This research is supported by the \textit{in-situ} measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: spectral modelling by means of the RTM Water Colour Simulator WASI,  underwater videometric measurements made by Olympus camera, aerial Google Earth and satellite images from different sources, spatial GIS and statistical analysis.

\subsubsection{General objective}\index{Research!objective!general}
The main objective of this study is to analyse the optical properties of the seagrass \textit{P.oceanica} and other seafloor types, and to apply the remote sensing techniques towards the investigation of the seagrass distribution in the selected locations along the northern coast of Crete.
General objectives:
	\begin{itemize}
		\item[*]Analysing spectral reflectance of the \textit{P.oceanica} and other seafloor cover types means of radiative transfer model tools (RTMs), using WASI.
		\item[*]Mapping spatial distribution of the seagrass \textit{P.oceanica} over selected locations along the northern coasts of Crete Island 
	\end{itemize}
